Loan Data Exploration by Xue Bai
========================================================
The data set contains 113,937 loans with 81 variables on each loan. I 
choose 10 varibles frome this data set to explore. The variables I choose are: 
loan original date, term, loan status, borrower rate, listing category, 
occupation, employment, monthly loan payment, loan original amount and 
income range. 

# Univariate Analysis
```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(dplyr)
library(GGally)
library(gridExtra)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
setwd("~/Downloads/udacity/project_4")
loan<-read.csv('prosperLoanData.csv',sep=',')
loan<-loan[,c('EmploymentStatus','Term','ListingCategory..numeric.',
              'Occupation','MonthlyLoanPayment','LoanOriginalAmount',
              'IncomeRange','CreditGrade','BorrowerState','LoanStatus',
              'BorrowerRate','LoanOriginationDate')]
loan$year<-as.numeric(substr(loan$LoanOriginationDate,1,4))
loan$year<-as.factor(loan$year)
loan$ListingCategory..numeric.<-as.factor(loan$ListingCategory..numeric.)
loan$Term<-as.factor(loan$Term)
head(loan)
```

# Univariate Plots Section
 
#### Number of loans in 2005-2014
 
```{r echo=FALSE,  message=FALSE, warning=FALSE, Univariate_Plots}
qplot(x=year,data=loan)
```

From this plot we can see that from 2005 to 2008, the number of loans increase. 
but at 2009, the number of loans suddenly decrease a lot. Then from 2009, the 
number of loans increase quiet fast and reach the highest at 2013. The dataset 
only contains data before 3/11/2014, so it is not surprise that the 
number of loans in 2014 is quiet small.
 
 
 
#### Term
 
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x=Term),data=loan)+geom_bar()
```

From this plot we can see that the most people choose to loan for 36 months.



The reason why people loan.

#### Listing Category
 
```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(aes(x=ListingCategory..numeric.),data=loan)+geom_bar()+ 
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=.5))+scale_x_discrete(limits=seq(0,22,1),breaks=1:21,labels=c(
    'Not Available','Debt Consolidation','Home Improvement',
    'Business', 'Personal Loan','Student Use','Auto','Other',
    'Baby&Adoption', 'Boat','Cosmetic Procedure', 'Engagement Ring',
    'Green Loans','Household Expenses', 'Large Purchases', 
    'Medical/Dental', 'Motorcycle','RV', 'Taxes', 'Vacation', 'Wedding Loans'))+labs(x='Listing Category',y='count')+
  scale_y_continuous(trans='log10') +coord_flip()
```

From this plot we can see that most people loan because of debt consolidation.



#### Loan Status

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(aes(x=LoanStatus),data=loan)+geom_bar() + 
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=.5))
summary(loan$LoanStatus)
```

Most of the loans are completed or in current. Few of loans past due.


#### BorrowerState

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(aes(x=BorrowerState),data=loan)+geom_bar()+
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=.5))
```

Form this plot we can find that the number of loans in CA is significantly 
large.
The number of loans in FL,IL,NY,TX is also very large.
 
 
 
#### Occupation
 
```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(aes(x=Occupation),data=loan)+geom_bar()+
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=.5))
```
 

#### Employment
 
```{r echo=FALSE,message=FALSE, warning=FALSE}
# remove the blank for empolyment Status
ggplot(aes(x=EmploymentStatus),data=subset(loan,EmploymentStatus!=''))+
  geom_bar()+
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=.5))
```


#### Incomerange
 
```{r echo=FALSE,message=FALSE, warning=FALSE}
loan$IncomeRange=factor(loan$IncomeRange,levels=c('Not employed',
'Not displayed','$0','$1-24,999','$25,000-49,999','$50,000-74,999',
                                '$75,000-99,999','$100,000+'))
ggplot(aes(x=IncomeRange),data=loan)+geom_bar()+
  theme(axis.text.x = element_text(angle=90, hjust=0, vjust=.5))
```

From this plot we can find that most people loan has income range between 
25,000 to 74,999.



#### Loan Amount
 
```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(aes(x=LoanOriginalAmount),data=loan)+geom_histogram()+
  scale_x_log10(breaks=c(1000,2500,5000,10000,20000,27000,40000))
```

Most people's loan original amount is below 27,000.



#### Credit Grade

```{r echo=FALSE,message=FALSE, warning=FALSE}
p1=ggplot(aes(x=CreditGrade),data=loan)+geom_bar()
# exclude blank one
p2=ggplot(aes(x=CreditGrade),data=subset(loan,CreditGrade!=""))+
  geom_bar()
grid.arrange(p1,p2)
```

Most of the credit grades are leave blank, we need to ignore the blank credit 
grades when making the Bivariate and mulivariate plot. 


#### Borrower Rate
 
```{r echo=FALSE,message=FALSE, warning=FALSE}

ggplot(aes(x=BorrowerRate),data=loan)+geom_histogram()+
  xlim(0,0.4)

```

Most of the borrower rate is between 0.05-0.35 which is a really large rate. 
The distribution of the rate is like normal distribution.






### What is the structure of your dataset?

The data set contains 113,937 loans with 81 variables on each loan. I choose 12 varibles frome this data set to explore. The variables I choose are: 
loan original date, term, loan status, borrower rate, listing category, 
occupation, employment, monthly loan payment, loan original amount,income range, credit grade and borrower state.

term:The length of the loan expressed in months.(12,36,60)
loan status:Cancelled, Chargedoff, Completed, Current, Defaulted, FinalPaymentInProgress, PastDue
listing category:The category of the listing that the borrower selected when 
posting their listing(including: Not Available, Debt Consolidation, Home 
Improvement,  Business, Personal Loan, Student Use, Auto, Other, Baby&Adoption, 
Boat, Cosmetic Procedure, Engagement Ring, Green Loans,  Household Expenses, 
Large Purchases, Medical/Dental, Motorcycle, RV, Taxes, Vacation, Wedding Loans

### What is/are the main feature(s) of interest in your dataset?

I am interest in what factor influence the rant of loans? And what factor is 
the most important? The factors I'm interested include: 1. when 2. where 
3.term of loan 4.occupation/employment status of borrower 5. loan amount 6. Incomerange of borrower7. credit grade of borrower

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

I also explore the reason why people loan. 

### Did you create any new variables from existing variables in the dataset?

I create year to indicate the year when te loan start.

### Of the features you investigated, were there any unusual distributions?
###Did you perform any operations on the data to tidy, adjust, or change the 
###form of the data? If so, why did you do this?

I change the Listing Category of loan to words instead of the number to make 
it easier for people to know what is going.

# Bivariate Plots Section

```{r fig.width=8, fig.height=6,echo=FALSE, message=FALSE, warning=FALSE,Bivariate_Plots}

set.seed(1234)
ggpairs(loan[sample.int(nrow(loan),1000),c('EmploymentStatus',
'Term','ListingCategory..numeric.','MonthlyLoanPayment',
'LoanOriginalAmount','IncomeRange','CreditGrade','BorrowerState',
'BorrowerRate','year')],
 axisLabels = "none",showStrips = FALSE,columnLabels=c('ES','Term',
'LC','MP','LOA','IR','CG','BS','BR','year'))+
  theme(axis.text=element_blank(),axis.line=element_blank(), 
axis.ticks=element_blank(),
axis.title.x=element_blank(),panel.grid.major= element_blank())

```

From the plot below we can see that the average borrower rate is different year 
to year. And there are only few outliers. The average borrower rate is quite 
small at 2015. From 2006 to 2008, the average borrower rate decrease a little 
bit. From 2008-2011, the borrower rate increase. And from 2011 to 2014, the 
average borrower rate decrease.

#### year-rate
 
```{r echo=FALSE,message=FALSE, warning=FALSE}
loan$year_num<-as.numeric(substr(loan$LoanOriginationDate,1,4))
loan$month_year<-
  as.numeric(substr(loan$LoanOriginationDate,1,4))+
  as.numeric(substr(loan$LoanOriginationDate,6,7))/12
ggplot(data=loan, aes(x=month_year, y=BorrowerRate))+
  geom_point(alpha=1/100)+
  stat_summary(aes(x=month_year),fun.y = 'mean',colour = 'red',geom='line')+scale_x_continuous(breaks=seq(2005,2014,1))


```





From this plot, we can find that the longer the term is, higher rate is likely.

#### Term-rate
 
```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(loan,aes(x=Term,y=BorrowerRate))+geom_boxplot()

```






The difference of loan rate to different employment status is not much. However,
the rate for borrower that is not employed is higher than other status.

#### Employment-rate
 
```{r echo=FALSE,message=FALSE, warning=FALSE}
term_rate<-loan %>%
  group_by(EmploymentStatus) %>%
  summarise(mean=mean(BorrowerRate),se=sd(BorrowerRate)/sqrt(n()))

ggplot(data=loan, aes(x=EmploymentStatus, y=BorrowerRate))+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=.5))

```





From the scatter plot I find the rate variance decrease as the loan original 
amount increase.But before 25,000 the loan original amount seems has litter 
relation to the loan original amount. For the loan over 25,000 dollars, the 
loan rate variance is apparently smaller. 

The blue line is the correlation line of borrower rate to loan original amount.
From this line, we can find that the larger the loan original amount is, the 
lower borrowerrate might be.
#### LoanAmount-rate
 
```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(data=loan, aes(x=LoanOriginalAmount, y=BorrowerRate))+
  geom_point()
ggplot(data=loan, aes(x=LoanOriginalAmount, y=BorrowerRate))+
  geom_point(alpha=1/50)+stat_smooth(method='lm')
```






The borrower rate has little relation to the borrower state.

#### borrowerstate-rate
 
```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(data=loan, aes(x=BorrowerState, y=BorrowerRate))+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=.5))
```






For borrower rate is lower when the income range of the borrower is increase. 
But surprisely, the income range of 0 is lowest. There maybe some special loan 
project for these people.

#### IncomeRange-rate
 
```{r echo=FALSE,message=FALSE, warning=FALSE}

ggplot(data=loan, aes(x=IncomeRange, y=BorrowerRate))+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=.5))
```







Apparently, the higher the credit grade, the lower rate is.

#### Credit Grade
 
```{r echo=FALSE,message=FALSE, warning=FALSE}
# ignore blank credit grade
ggplot(data=subset(loan,CreditGrade!=''), aes(x=CreditGrade, y=BorrowerRate))+geom_boxplot()+
  scale_x_discrete(limits=c('AA','A','B','C','D','E','HR','NC'))
```




#### The relation between category of loan to loan original amount

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(data=loan, aes(x=ListingCategory..numeric., y=LoanOriginalAmount))+geom_boxplot()+coord_flip()
```






# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in 
###the dataset?

Some factors I think may have impact on the loan rate actually has little 
impact on that. The factors I find that have relatively strong impact on loan 
rate are year of loan, term, income range and credit grade. 

### Did you observe any interesting relationships between the other features 
###(not the main feature(s) of interest)?

The loan original amount, the borrower state and the employment status has 
little relation to the loan rate.

### What was the strongest relationship you found?

The time of loan is the most important factor of borrower rate.


# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}
# get the loan original date, change the date only include year to include 
#year and month. The month is expressed in decimal part.
```



From the plot below we can see that before 2009, the term of loan is always 36 
months. After 2009, there are loans for 12, 26 and 60 months. When 12 and 36 
months' loan first appear(at 2010), the average rate is quiet low accordint to 
the rate for 36 months' loan. But it increase quiet fast. 

#### time-rate-term
 
```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(data=loan, aes(x=year, y=BorrowerRate))+
  geom_boxplot(aes(color=Term))

```






After 2009, no credit grade is available. But we can tell from 2005-2009, 
in general, the higher the credit grade is, the lower the average borrower rate.

#### time-rate-credit
 
```{r echo=FALSE,message=FALSE, warning=FALSE}
average_rate_year_credit<-  subset(loan,CreditGrade!='') %>%
group_by(CreditGrade,year) %>%
  summarise(average=mean(BorrowerRate))%>%
  arrange(year)%>%
  ungroup()

average_rate_year_credit$year<-
  as.numeric(average_rate_year_credit$year)+2004
ggplot(data=average_rate_year_credit,aes(x=year,y=average))+
  geom_line(aes(color=CreditGrade))+
  scale_color_brewer(limits=c('AA','A','B','C','D','E','HR','NC'))
```






The plot below descibe the relation of the average borrower rate in 2005-2014 
for people have different income range. I omit the 'Not displayed' category of 
the income range. From the below plot we can see that in genaral the higher the 
income range, thelower the average borrower rate is. And the basic shape of the average borrower rate to the year is the same. 

#### time-rate-Incomerange
 
```{r echo=FALSE,message=FALSE, warning=FALSE}
average_rate_year_Incomerange<-  loan %>%
group_by(IncomeRange,year) %>%
  summarise(average=mean(BorrowerRate))%>%
  arrange(year) %>%
ungroup()
average_rate_year_Incomerange$year<-
  as.numeric(average_rate_year_Incomerange$year)+2004

average_rate_year_Incomerange_sub<-subset(
  average_rate_year_Incomerange,IncomeRange!='Not displayed')
ggplot(average_rate_year_Incomerange_sub,aes(x=year,y=average))+
  geom_line(aes(color=IncomeRange))+
  scale_x_continuous(breaks=seq(2005,2014,1))+
  geom_point(aes(color=IncomeRange))+scale_color_brewer(
      limits=c('Not employed','$0','$1-24,999','$25,000-49,999',
               '$50,000-74,999','$75,000-99,999', '$100,000+'))

```


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of 
###looking at your feature(s) of interest?

The tendency of average loan rate is nearly the same for different terms, 
different Incomerange and credit grade. 

### Were there any interesting or surprising interactions between features?

At 2012-2014, the average loan rate decrease, but the average rate for 
unemployed people increase a lot.

------

# Final Plots and Summary

### Plot One

```{r echo=FALSE, Plot_One}
ggplot(aes(x=ListingCategory..numeric.),data=loan)+
  geom_bar()+ 
  theme(axis.text.x = element_text(angle=90,hjust=1,vjust=.5))+
  scale_x_discrete(limits=seq(0,21,1),breaks=1:21,
labels=c('Not Available','Debt Consolidation','Home Improvement',
'Business', 'Personal Loan','Student Use','Auto','Other',
'Baby&Adoption', 'Boat',  'Cosmetic Procedure', 'Engagement Ring',
'Green Loans','Household Expenses', 'Large Purchases', 'Medical/Dental', 'Motorcycle','RV', 'Taxes', 'Vacation', 'Wedding Loans'))+
  labs(x='Listing Category',y='count')+
  scale_y_continuous(trans='sqrt')+
  ggtitle('Loan Category')+
  labs(x='Loan Catagory',y='Number of loans')+coord_flip()
summary(loan$ListingCategory..numeric.)
```
### Description One

This plot describes number of loans for different categories. From this plot we 
can see that the most people loan for Debt Consolidation.

### Plot Two

```{r echo=FALSE, Plot_Two}
ggplot(data=loan, aes(x=month_year, y=BorrowerRate))+
  geom_point(alpha=1/100)+
  stat_summary(aes(x=month_year),fun.y ='mean',colour='red',geom='line')+
  scale_x_continuous(limits=
c(2005,2014.5),breaks=seq(2005,2014,1))+labs(x='Loan original year',
y='Loan Rate')+ggtitle('The loan rate in 2005-2014')
```


### Description Two

This plot shows the loan rate changes in 2005-2014. The red line is the average 
loan rate for that time. We can see from the plot that the loan rate increase in
2005-2011, and decrease from 2011 to 2014.

### Plot Three

```{r echo=FALSE, Plot_Three}
ggplot(average_rate_year_Incomerange_sub,aes(x=year,y=average))+
  geom_line(aes(color=IncomeRange))+
  scale_x_continuous(breaks=seq(2005,2014,1))+
  geom_point(aes(color=IncomeRange))+
  scale_color_brewer(limits=c('Not employed','$0','$1-24,999',
'$25,000-49,999','$50,000-74,999','$75,000-99,999', '$100,000+'))+
  ggtitle('Average borrower rate in 2005-2014 
          for different income range')+
  labs(x='Loan original year',y='Average Loan Rate')+
  ggtitle('The loan rate in 2005-2014 for different income range')
```


### Description Three

This plot descibe the relation of the average borrower rate in 2005-2014 
for people have different income range. I remove the 'Not displayed' category of 
the income range because it doesn't contain useful information. From the below 
plot we can see that in genaral the higher the income range, thelower the 
average borrower rate is. And the basic shape of the average borrower rate to
the year is the same.

------

# Reflection

The data set contains 113,937 loans with 81 variables on each loan. I choose 13 varibles frome this data set to explore its relation to the loan rate. The 
variables I choose are: loan original date, term, loan status, borrower rate,
listing category, occupation, employment, monthly loan payment, loan original 
amount, income range and borrower state. 

I started by understanding the individual variables in the data set, and then I explored the relation of each factor to the loan rate. Then I found the factors
which have may relate relatively strong to the loan rate and did some further explorement in these factors.

From the explorement, I find that the relatively strong factors of loan rate 
are date of loan, credit grade, income range of borrower and term. What 
surprise me is that the loan amount has little relation to the loan rate.

When doing this project, I find it is hard because many data is blank. 
For example, the credit grade is only available in 2005-2009 and the many 
employment status is not displayed. So I have to omit the this data when analyze
the data.

I also find that the loan rate is affected by too many features that I can't 
describe them all in one plot. So I need to split the features and find out the 
most important features or the features I am interested in and then display them
in different plots.

In this project, we know that the time, credit grade, income range and term has 
relatively strong relationships with the loan rate. In the further work, maybe
I can give each feature a coefficient which can makes us to predict the loan 
rate.